---
title: "ARCH and GARCH models"
author: "Taylor R. Brown"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(xts)
spyReturns <- diff(log(read.csv.zoo("data/some_factors.csv")$SPY.Adjusted))*100
xlkReturns <- diff(log(read.csv.zoo("data/spdrs.csv")$XLK.Adjusted))*100
xlkSpy <- merge(spyReturns, xlkReturns)
```


## Definitions

  - $P_t$: the closing price on day/week/month $t$ of stock/stock index
  - $X_t = \log P_t$: log-price...observed paths look very much like those of a random walk 
  - $Z_t = \triangledown X_t$: the log return
  - $100 Z_t$: percentage returns
  - $h_t = \text{Var}(Z_t|Z_{1:t-1})$ : the conditional variance aka volatility


## Motivation

```{r, out.width="80%"}
plot.ts(spyReturns)
```


## Motivation

<!-- TODO, homework question: Also, if we fit some ARMA model, $h_t$ is independent of $t$ and independent of $Z_{1:t-1}$\footnote{homework question}. This assumption is often violated. -->
<!-- \newline -->

We would like models that take into account *stylized features* that appear in real-world data such as

  - **tail-heaviness**
  - **asymmetry**
  - **volatility clustering** 
  - **serial dependence without correlation** 
  
We introduce AutoRegressive Conditional Heteroscedasticity (ARCH) models, Generalized AutoRegressive Conditional Heteroscedasticity (GARCH), and stochastic volatility (SVOL) models in this slide deck.




## Definition: ARCH(p)

\begin{block}{ARCH(p) (Engle 1982)}
\begin{align*}
Z_t &= \sqrt{h_t}e_t, \hspace{10mm} \{e_t\} \sim IID(0,1) \\
h_t &= \alpha_0 + \sum_{i=1}^p \alpha_i Z^2_{t-i}.
\end{align*}
$\alpha_0 > 0$, $a_i \ge 0$, $p \in \mathbb{N}$
\end{block}

\begin{itemize}
\item volatility increases if we have observed big movements
\item setting $a_i$ to $0$ gives us white noise model
\item $\{e_t\}$ is sometimes but not always assumed to be Normal
\end{itemize}


## Definition: GARCH(p,q) model

\begin{block}{GARCH(p,q)}
\begin{align*}
Z_t &= \sqrt{h_t}e_t, \hspace{10mm} \{e_t\} \sim IID(0,1) \\
h_t &= \alpha_0 + \sum_{i=1}^p \alpha_i Z^2_{t-i} + \sum_{j=1}^q \beta_j h_{t-j}.
\end{align*}
$\alpha_0 > 0$, $a_i \ge 0$, $\beta_j \ge 0$, $p \in \mathbb{N}$
\end{block}

\begin{itemize}
\item now volatility is a function of its own past values, in addition to the past observations
\item $\{e_t\}$ may or may not be normal
\end{itemize}

## Definition: Stochastic Volatility model


\begin{block}{Stochastic Volatility Model}
\begin{align*}
Z_t &= \sqrt{h_t}e_t, \hspace{10mm} \{e_t\} \sim IID(0,1) \\
\ln h_t &= \gamma_0 + \gamma_1 \ln h_{t-1} + \eta_t, \hspace{10mm} \{\eta_t\} \sim IID(0,\sigma^2). 
\end{align*}
where $\{\eta_t\}$ and $\{e_t\}$ are independent.
\end{block}

\begin{itemize}
\item log-volatility is an AR(1) process. 
\item $\gamma_1$ is usually around $.95$.
\item more difficult to estimate (intractable likelihood)
\end{itemize}



## ARCH(1)

In the case of ARCH(1), $h_t = \alpha_0 + \alpha_1 Z_{t-1}^2$ and

\begin{align*}
Z_t^2 &= h_t e_t^2\\
&= [\alpha_0 + \alpha_1 Z^2_{t-1}] e_t^2 \\
&= \alpha_0 e_t^2 + \alpha_1 h_{t-1} e_{t-1}^2 e_t^2 \\
&= \alpha_0 e_t^2 + \alpha_1 [\alpha_0 + \alpha_1 Z_{t-2}^2] e_{t-1}^2 e_t^2 \\
&= \alpha_0 e_t^2 + \alpha_1 \alpha_0e_{t-1}^2 e_t^2 + \alpha_1^2 Z_{t-2}^2 e_{t-1}^2 e_t^2 \\
&= \alpha_0 e_t^2 + \alpha_1 \alpha_0e_{t-1}^2 e_t^2 + \alpha_1^2 [h_{t-2}e_{t-2}^2] e_{t-1}^2 e_t^2 \\
&= \alpha_0 e_t^2 + \alpha_1 \alpha_0e_{t-1}^2 e_t^2 + \alpha_1^2 [\alpha_0 + \alpha_1 Z_{t-3}^2 ] e_{t-2}^2 e_{t-1}^2 e_t^2 \\
&=  \left\{\alpha_0 e_t^2 + \alpha_1 \alpha_0e_{t-1}^2 e_t^2  + \alpha_1^2 \alpha_0 e_{t-2}^2 e_{t-1}^2 e_t^2\right\} + \left\{\alpha_1^3  Z_{t-3}^2 e_{t-2}^2 e_{t-1}^2 e_t^2 \right\} \\
&= \alpha_0 \sum_{j=0}^n \alpha_1^j e_t^2 e_{t-1}^2 \cdots e_{t-j}^2 + a_1^{n+1}Z_{t-n-1}^2 e_t^2 e_{t-1}^2 \cdots e_{t-n}^2
\end{align*}


## ARCH(1)

\[
Z_t^2 = \left( \alpha_0 \sum_{j=0}^n \alpha_1^j e_t^2 e_{t-1}^2 \cdots e_{t-j}^2\right) + \left(a_1^{n+1}Z_{t-n-1}^2 e_t^2 e_{t-1}^2 \cdots e_{t-n}^2\right)
\]
If $\alpha_1 < 1$:
\begin{itemize}
\item second term goes to $0$ as $n \to \infty$
\item first term has a limit, let's call it $\alpha_0 \sum_{j=0}^{\infty}\alpha_1^j (e_t^2 \times \cdots \times e_{t-j}^2)$
\end{itemize}
so
\begin{block}{}
\[
Z_t^2 = \alpha_0 \sum_{j=0}^{\infty} \alpha_1^j e_t^2 e_{t-1}^2 \cdots e_{t-j}^2
\]
\end{block}
if we're looking at an infinitely long sequence.


## ARCH(1)


Weakly-stationary!

\[
E[Z_t] = E[\sqrt{h_t}e_t] = E[\sqrt{h_t}]E[e_t] = 0
\]
Marginal variance
\begin{align*}
\text{Var}[Z_t] &= E[Z_t^2] \\
&= E[\alpha_0 \sum_{j=0}^{\infty} \alpha_1^j e_t^2 e_{t-1}^2 \cdots e_{t-j}^2] \tag{previous slide}\\
&= \alpha_0 \sum_{j=0}^{\infty} \alpha_1^j = \alpha_0/(1-\alpha_1) \tag{linearity, independence, geom. series}
\end{align*}
Autocovariance
\begin{align*}
\gamma_Z(h) &= E[Z_{t+h}Z_t] = E[E(Z_{t+h}Z_t| e_s, s < t+h)] \tag{LTE} \\
&= E[Z_t E(Z_{t+h}| e_s, s < t+h)] = 0
\end{align*}

But remember that volatility in this case is the **conditional** variance. 

## ARCH(1)


The ARCH(1) model is white noise, but not IID noise. 

\begin{align*}
E[Z_t^2|Z_{1:t-1}] &= E[(\alpha_0 + \alpha_1 Z_{t-1}^2)e_t^2|Z_{1:t-1}] \\
&= (\alpha_0 + \alpha_1 Z_{t-1}^2) E[e_t^2|Z_{1:t-1}]\\
&= (\alpha_0 + \alpha_1 Z_{t-1}^2) \\
\end{align*}

This depends on $Z_{t-1}$. It is **not** $E[Z_t^2] = \alpha_0 / (1- \alpha_1)$.

## ARCH(1)


$$
E[Z_t^2|Z_{1:t-1}] = (\alpha_0 + \alpha_1 Z_{t-1}^2)
$$
So
\begin{align*}
\gamma(1)_{Z^2} 
&= \mathbb{E}[Z_t^2 Z_{t-1}^2] \\
&= E\left[Z_{t-1}^2 E\left(Z_t^2 \mid Z_{1:t-1} \right) \right] \\
&= E\left[ Z_{t-1}^2(\alpha_0 + \alpha_1 Z_{t-1}^2) \right] \\
&= \alpha_0 + \alpha_1 \gamma(0)_{Z^2}
\end{align*}

This is why we look at autocorrelation of the squared return process.

TODO: homework generalize to lag greater than 1

<!-- ## ARCH(1) -->


<!-- Let's write down the conditional likelihood: -->
<!-- \begin{align*} -->
<!-- L &= f(z_n, z_{n-1},\ldots,z_2 |z_1) \\ -->
<!-- &= \prod_{t=2}^n f(z_t|z_{1:t-1}) \\ -->
<!-- &= \prod_{t=2}^n \frac{1}{\sqrt{2 \pi h_t } } \exp\left[- \frac{z_t^2}{2 h_t } \right] \tag{cndtl nrmlty} \\  -->
<!-- &= \prod_{t=2}^n \frac{1}{\sqrt{2 \pi (\alpha_0 + \alpha_1 z_{t-1}^2) } } \exp\left[- \frac{z_t^2}{2(\alpha_0 + \alpha_1 z_{t-1}^2)  } \right] \tag{defn. $h_t$} -->
<!-- \end{align*} -->

TODO this can be made into a hw question ^

## Example 1

Fake data or real data?

```{r, echo=F, eval=T}
suppressMessages(library(fGarch, quietly = T, warn.conflicts = F))
spec <- garchSpec(model = list(omega = 1, alpha = c(0.5),
                  beta = 0))
zt <- garchSim(spec, n = 1000)
par(mfrow=c(3,1))
plot.ts(zt)
acf(zt)
acf(zt^2)
par(mfrow=c(1,1))
```



## Example 1

Fake!

```{r, echo=TRUE, eval=FALSE}
library(fGarch)
spec <- garchSpec(model = list(omega = 1, alpha = c(0.5),
                  beta = 0))
zt <- garchSim(spec, n = 1000)
```




## Example 2

Another example ARMA(1,1) + GARCH(1,1): 
\[
X_t - \mu = \phi(X_{t-1} - \mu) + Z_t + \theta Z_{t-1} 
\]
\[
Z_t = \sqrt{h_{t}}e_t
\]
\[
h_t = \alpha_0 + \alpha_1 Z_{t-1}^2 + \beta_1 h_{t-1}
\]

## Example 2

\includegraphics[width=75mm]{pics/garch/Rplot11.png}
\includegraphics[width=575mm]{pics/garch/Rplot12.png}


See code for more examples on how to simulate and estimate ARCH/GARCH models.



## Sources:

Chapters 7.1,7.2 of Introduction to Time Series and Forecasting Brockwell/Davis